Different forms of uncertainty differently affect the evolution of social learning

### 2022-08-11 ms.tex
Updates from Overleaf

\begin{abstract} 
  Social learning is essential to human survival---it is a critical
  adaptation for dealing with uncertainty. However, uncertainty takes many forms.
  We identified four theoretically important types of uncertainty: temporal
  environmental variability, ambiguous payoffs, decision set size, and effective
  lifespans. We used this framework to study conditions for the evolution of
  social learning in populations with cognitively realistic individual learning
  capacities.  We used evolutionary agent-based modeling to track how often social
  learning evolved. Agents performed one of several behaviors to acquire payoffs.
  Individual learning used the softmax algorithm. Social learning occurs between
  generations  through vertical and oblique transmission.  Different types of
  uncertainty had different effects. Temporal environmental variability suppressed
  social learning. Larger decision set sizes promoted social learning even
  when environmental change was more likely than not. Payoff
  ambiguity and lifespan had more complex effects on social learning, dependent on
  other uncertainty parameters.  This shows the importance of clearly
  operationalizing uncertainty in theories of social learning evolution.
% Social learning is essential to human survival. It is likely to evolve when it is more
% efficient than asocial, trial-and-error learning. Theoretically, social learning
% is adaptive for some uncertainty, but too much uncertainty makes
% social information unreliable. This fact is empirically
% supported across biology and human sciences. However, it is unclear how
% specific classes and types of uncertainty affect social
% learning evolution. Furthermore, existing models and experimental operationalizations of
% uncertainty are ambiguously related, and 
% tend to only consider a small number of behavioral
% choices.  Here we use evolutionary agent-based modeling to consolidate
% models of uncertainty in social learning evolution to address these issues.
% We model societies of agents who perform one of a number
% of possible behaviors to acquire payoffs in a time-varying environment.
% We identified complex patterns in social learning evolution depending on contextual
% uncertainty variables: environmental variability, ambiguous payoffs, 
% decision sets size, and effective lifespans. 
% Our work advances social learning evolution theory in a way that could help 
% guide human, non-human, and artificial intelligence towards optimal responses 
% to existential threats and new opportunities.
\footnote{This document contains
\total{citnum} references.}  
\end{abstract}

@@

Other (asocial) forms of learning may also be adaptations to uncertainty that could be
accounted for, but are typically glossed over.  In general, learning is an iterative
process by which one repeatedly acquires information, forms representations and
predictions, and tests and refines those representations and predictions to manage
uncertainty~\cite{jacobs2011bayesian,clark2013whatever}.  Because cultural evolutionary
models are more often used to explain social learning mechanisms, they tend to use a
single, weak individual-learning mechanism for comparison. Abstracting away asocial forms
of learning may lead us to underestimate the extent to which individual learning provides
quality information for social learning and for the individual to adapt to the challenges
assumed to be solved by social learning. 
% underestimate the quality of information
% available for social learning.

To address these concerns we developed an evolutionary agent-based model that
simultaneously operationalizes several different forms of uncertainty and endows agents
with a relatively powerful learning mechanism. We have chosen to model four kinds of
uncertainty that are common in cultural evolution models and the related empirical
literature:  (1) temporal environmental variability; (2) selection set size (the number of
possible behaviors); (3) payoff ambiguity (the difference in the expected rewards for
different behaviors); and (4) the effective lifespan of agents (the number of
opportunities for individual learning), explained below. The model also assumes a more
cognitively realistic and empirically-motivated learning mechanism, the softmax algorithm.
This allows for a more accurate assessment of tradeoffs between 
social and asocial learning
by fleshing out an individual-level learning mechanism that makes full use of 
the individually and socially learned data. 
In this work, social learners learn from the previous generation, and 
asocial learners do not. All agents engage in individual-level learning within
their lifespans whether they are social learners or not.
We use this integrated model to ask what kinds of uncertainty are likely to
favor social learning, and to clarify the logic of why this would be the case.

@@

\textbf{Temporal environmental variability.} When external or internal shocks
(e.g., climatic events, migration, technological paradigm shifts) occur,
strategies that were previously adaptive may no longer be optimal. The difficulty
in predicting either when such shocks will occur or what behaviors will be
adaptive in the resulting environments leads to uncertainty.

>>>>

\begin{abstract} 
  Social learning is essential to human survival---it is a critical
  adaptation for dealing with uncertainty. However, uncertainty takes many forms.
  We identified four theoretically important types of uncertainty: temporal
  environmental variability, ambiguous payoffs, decision set size, and effective
  lifespans. We used this framework to study conditions for the evolution of
  social learning in populations with cognitively realistic individual learning
  capacities.  We used evolutionary agent-based modeling to track how often social, rather than individual,
  learning evolved. Agents performed one of several behaviors to acquire payoffs.
  Individual learning used the softmax algorithm. Social learning occurs between
  generations  through vertical and oblique transmission.  Different types of
  uncertainty had different effects. Temporal environmental variability suppressed
  social learning. Larger decision set sizes promoted social learning even
  when the environment changed frequently. Payoff
  ambiguity and lifespan had more complex effects on social learning, dependent on
  other uncertainty parameters.  This shows the importance of clearly
  operationalizing uncertainty in theories of social learning evolution.
% Social learning is essential to human survival. It is likely to evolve when it is more
% efficient than asocial, trial-and-error learning. Theoretically, social learning
% is adaptive for some uncertainty, but too much uncertainty makes
% social information unreliable. This fact is empirically
% supported across biology and human sciences. However, it is unclear how
% specific classes and types of uncertainty affect social
% learning evolution. Furthermore, existing models and experimental operationalizations of
% uncertainty are ambiguously related, and 
% tend to only consider a small number of behavioral
% choices.  Here we use evolutionary agent-based modeling to consolidate
% models of uncertainty in social learning evolution to address these issues.
% We model societies of agents who perform one of a number
% of possible behaviors to acquire payoffs in a time-varying environment.
% We identified complex patterns in social learning evolution depending on contextual
% uncertainty variables: environmental variability, ambiguous payoffs, 
% decision sets size, and effective lifespans. 
% Our work advances social learning evolution theory in a way that could help 
% guide human, non-human, and artificial intelligence towards optimal responses 
% to existential threats and new opportunities.
\footnote{This document contains
\total{citnum} references.}  
\end{abstract}

@@

Other (asocial) forms of learning are also likely adaptations to uncertainty that should be
accounted for, but are typically glossed over in models.  In general, learning is an iterative
process by which one repeatedly acquires information, forms representations and
predictions, and tests and refines those representations and predictions to manage
uncertainty~\cite{jacobs2011bayesian,clark2013whatever}.  Because cultural evolutionary
models are more often used to explain social learning mechanisms, they tend to use a
single, weak individual-learning mechanism for comparison. Such modelling choices may lead us to underestimate the extent to which individual learning provides
quality information, both for asocially adapting to uncertainty and for social learners to take advantage of.
%and for the individual to adapt to the challenges
assumed to be solved by social learning. 
% underestimate the quality of information
% available for social learning.

To address these concerns we developed an evolutionary agent-based model that
simultaneously operationalizes several forms of uncertainty and endows agents
with a relatively powerful learning mechanism. We model four kinds of
uncertainty that are common in cultural evolution models and the related empirical
literature:  (1) temporal environmental variability; (2) selection set size (the number of
possible behaviors); (3) payoff ambiguity (the difference in the expected rewards for
different behaviors); and (4) the effective lifespan of agents (the number of
opportunities for individual learning), explained below. The model also assumes a more
cognitively realistic and empirically-motivated learning mechanism, the softmax algorithm.
This allows for a more accurate assessment of tradeoffs between 
social and asocial learning
by fleshing out an individual-level learning mechanism that makes full use of 
the individually and socially learned data. 
In this work, social learners learn from the previous generation, and 
asocial learners do not. All agents engage in individual-level learning within
their lifespans whether they are social learners or not.
We use this integrated model to ask which kinds of uncertainty are likely to
favor social learning and to clarify the logic underlying the results.
%of why this would be the case.

@@

\textbf{Temporal environmental variability.} When environments change
(e.g., climatic events, migration, technological paradigm shifts),
strategies that were previously adaptive may no longer be optimal. The difficulty
in predicting either when such shocks will occur or which behaviors will be
adaptive in the resulting environments leads to uncertainty.
